---
title: "Coursera Reproducible Research: Program Assignment 1"
author: "ncapps"
date: "March 8, 2017"
output: html_document
---

## Loading and preprocessing the data
1. Load the data
2. Process/transform the data
```{r echo=TRUE, message=FALSE}
library(lubridate)
library(dplyr)
library(ggplot2)
library(mice)
filename <- "activity.csv"
if (!file.exists(filename)) { 
        unzip("activity.zip") # unzip file data
}

df <- read.csv(filename, header=TRUE, na.strings="NA", sep=",", stringsAsFactors = FALSE)
df$date <- ymd(df$date) # convert to date datatype
```



## What is the mean total number of steps taken per day?
1. Calculate the total number of steps taken per day
2. Make a histogram of the total number of steps taken each day
3. Calculate the mean and median of the total number of steps taken per day
```{r echo=TRUE, fig.path="instructions_fig/plot1"}
sumsteps <- df %>% 
        filter(!is.na(steps)) %>% ## ignore missing values
        group_by(date) %>% 
        summarise(steps = sum(steps)) ## total steps per day

qplot(date, data=sumsteps, geom="histogram", weight=steps, ylab="Steps", binwidth=1)

summary(sumsteps$steps) ## Median / Mean to total steps per day
```



## What is the average daily activity pattern?
1. Make a time series plot of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)
2. Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?
```{r echo=TRUE, message=FALSE}
avgsteps <- df %>% 
        filter(!is.na(steps)) %>% ## ignore missing values
        group_by(interval) %>% 
        summarise(steps = mean(steps)) %>% ## avg steps per interval
        arrange(desc(steps)) ## sort by steps descending

qplot(interval, steps, data=avgsteps, geom="line")

head(avgsteps,1) ## interval 835 averages the most steps
```

## Imputing missing values
1. Calculate and report the total number of missing values in the dataset
2. Fill missing values with mean for that 5-minute interval
3. Create a new dataset that is equal to the original dataset but with the missing data filled in.
4. Make a histogram of the total number of steps taken each day
5. Calculate and report the mean and median total number of steps taken per day. 
6. Do these values differ from the estimates from the first part of the assignment? What is the impact of imputing missing data on the estimates of the total daily number of steps?

```{r echo=TRUE,message=FALSE}
df2 <- df
md.pattern(df2) ## 2304 rows with NAs


df2[is.na(df2),"steps"] <- ## impute missing steps with interval mean
        do.call(rbind, sapply(df2[is.na(df2),"interval"], function(x){
        avgsteps[avgsteps$interval==x,"steps"]}))

sumsteps2 <- df2 %>% 
        group_by(date) %>% 
        summarise(steps = sum(steps)) ## total steps per day

qplot(date, data=sumsteps2, geom="histogram", weight=steps, ylab="Steps", binwidth=1)

summary(sumsteps2$steps)
```
The imputing missing data with the mean of a given interval increased the median total steps per day by 10. The mean total steps per day did not change.

## Are there differences in activity patterns between weekdays and weekends?
1. Create a new factor variable in the dataset with two levels - "weekday" and "weekend" indicating whether a given date is a weekday or weekend day.
2. Make a panel plot containing a time series plot of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis).

```{r echo=TRUE, message=FALSE}
df2$weekday <- sapply(df2$date, function(x){
        if(weekdays(x) %in% c("Saturday","Sunday")){
                "weekend"
        }else{
                "weekday"
        }}) ## Create weekend/weekday factor variable

avgsteps2 <- df2 %>% 
        group_by(interval,weekday) %>% 
        summarise(steps = mean(steps)) ## avg steps per interval / weekday

qplot(interval,steps, data=avgsteps2, geom="line", facets=weekday~.)
```

